Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-10-15 crowdsource/article.tex
Improve text

\usepackage{algorithmic}

@@

\emph{
In the Results section "A large variety of performing methods", we intend to
assemble together various layman presentations of different approaches. If you
could provide a short popular-science style description of your approach (and
specifically of what you consider their main characterizing ingredients,
INVERSE COVARIANCE for you) this would be very helpful for me.  Here is not
really an issue of being precise, but of commenting substantial elements to a
public of... ignorant biologists! What's the logic of making a certain step?
How does the method work in a nutshell, without using equations but expressed
in plain words, etc. ? Then, based on all the texts I will collect from you, I
would build the section and let you edit it to correct mistakes or add
precisions.\\
}

\paragraph{A first version}
The inference problem reduced to its simpliest shape is the identification of
neuron pairs that are directly connected in the network. Intuitively, such
pairs should be related somehow and in particular their activities have to be
similar since their are dependent on the connectivity network. Therefore a way
to identify them is to measure, or at least estimate, how closely related two
neurons are regarding their activities. The most common measure for that
purpose is the (Pearson) correlation.  However, correlation also measures the
degree of association between neurons that are connected through others and so 
connected indirectly. It means all related neurons are found but direct pairs
are indinstinguishable from indirect ones.

A variant of this classic approach is the partial correlation which measures
the degree of association between neurons but takes into account the state
(here, the activity state) of other neurons. If all other neurons are
considered, the partial correlation is known as the inverse covariance. By
conditioning on them (and under a few assumptions), inverse correlation only
detects direct association between neuron pairs and filter out indirect
relationships that may also be observed in data \citep{sutera2014simple}. The
interest of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks \citep{de2004discovery,Schafer:2005}.

Nevertheless, in particular for the connectome inference from fluorescent
data, a pre-processing is required before applying either of both measures.
Such processes (i) abstract the activity signal from fluorescent data; and
(ii) highlight characteristics of activity signals (e.g., spikes) to make
easier the assessement of the degree of association and therefore improve the
quality of the estimates of both measures.

Accordingly, ahead of (partial) correlation, filters are applied on
fluorescent data in order to (i) smoothe the signal\footnote{I assume that
noise will have been explained at this stage of the paper. Otherwise, we
should mention something like this :\textit{Noise is due to fluctuations
independent of calcium, calcium fluctuations independent of spiking activity,
calcium fluctuations in nearby tissues that have been mistakenly captured, or
simply by the imaging process.}}; (ii) emphasise short spikes that are
indirect indicator of neuron communication; and (iii) artificially magnify
spikes that occur during normal activity because we have conjectured that when
a large part of the network is firing (i.e., called a network burst
\citep{stetter2012model}), the estimation of degree of assocation is
meaningless \citep{sutera2014simple}.

\paragraph{Alternative}
A simple yet effective solution to the problem of connectome inference in
calcium imaging data consists of two steps. \cite{sutera2014simple} First,
processing the raw signals and to detect neural peak activities while filtering
noise. Second, inferring the degree of association between neurons from partial
correlation statistics, the inverse of the correlation matrix. This statistical
measure has the property to filter indirect associations between neuron and
gives a high degree of association between directly connected neurons. This
approach is the winning methods of the Connectomics challenge.

@@

The method involving inverse covariance, fully described in \cite{sutera2014simple}, is presented in Table~\ref{alg:inverse-covariance-pseudocode}. Under the simplifying assumption that neurons are on-off units defined by short intense periods of activity (i.e., spikes) and long periods of inactivity, it is possible to extract that information from the fluorescent data by appling several filters, common in the field of signal processing \citep{kaiser1977data, oppenheim1983signals}.

The first step of our pipeline is to smoothe the fluorescence signal by applying a low-pass filter for filtering out high frequency noise:
\begin{align}
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\label{eq:weighted-asymetric-median}
\end{align}

@@

The objective of the last step of our filtering procedure is to decrease the importance of spikes that occur when there is high global activity in the network with respect to spikes that occur during normal activity. Indeed, we have conjectured that when a large part of the network is firing, the rate at which observations are made is not high enough to be able to detect interactions, and that it would therefore be preferable to lower their importance by changing their magnitude appropriately. Additionally, it is well-known that neurons may also spike because of a high global activity \citep{stetter2012model}. In such context, detecting pairwise neuron interactions from the firing activity is meaningless. As such, the signal output by h is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}, \label{eq:magnify-filter}
\end{align}
whose effect is to magnify the importance of spikes that occur in cases of low global activity (measured by $\sum_j x^t_j$).\\

Under the assumption that fluorescence level neuron are jointly gaussian, \textit{partial correlation} thus measures conditional dependencies between variables ; therefore it should naturally only detect direct associations between neurons and filter out spurious indirect effects as mentioned before.

@@

\begin{table}[ht]
\begin{algorithmic}
 \algsetup{linenodelimiter= }

@@

\end{algorithmic}
\caption{Inverse covariance pseudo-code}
\label{alg:inverse-covariance-pseudocode}
\end{table}

>>>>

\usepackage{algorithm}
\usepackage{algorithmic}

@@

\emph{
In the Results section "A large variety of performing methods", we intend to
assemble together various layman presentations of different approaches. If you
could provide a short popular-science style description of your approach (and
specifically of what you consider their main characterizing ingredients,
INVERSE COVARIANCE for you) this would be very helpful for me.  Here is not
really an issue of being precise, but of commenting substantial elements to a
public of... ignorant biologists! What's the logic of making a certain step?
How does the method work in a nutshell, without using equations but expressed
in plain words, etc. ? Then, based on all the texts I will collect from you, I
would build the section and let you edit it to correct mistakes or add
precisions.\\}

A partial correlation approach \cite{sutera2014simple}  is a simple yet effective
solution to the problem of connectome inference in calcium imaging data.
This winning approach of the Connectomics challenge consists of two steps.
First, time-series are processed using standard
signal processing filters in order to : (i) remove noise mainly due to
fluctuations independent of calcium, calcium fluctuations independent of
spiking activity, calcium fluctuations in nearby tissues that have been
mistakenly captured, or simply by the imaging process ; (ii) to account for
fluorescence low decay ; and (iii) to reduce the importance of high global
activity in the network. Secondly, inferring the degree of association between
neurons is then made with partial correlation statistics, the inverse of the
correlation matrix. This statistical measure has the property to filter
indirect associations between neurons and gives a high degree of association
between directly connected neurons.

@@

The inference problem reduced to its simplest shape is the identification of
neuron pairs that are directly connected in the network. Intuitively, such
pairs should be related somehow and in particular their activities have to be
similar since their are dependent on the connectivity network. Therefore a way
to identify them is to measure, or at least estimate, how closely related two
neurons are regarding their activities. The most common measure for that
purpose is the (Pearson) correlation.  However, correlation also measures the
degree of association between neurons that are connected through others and so
connected indirectly. It means all related neurons are found but direct pairs
are indistinguishable from indirect ones.

A variant of this classic approach is the partial correlation
\citep{sutera2014simple}, also known as the inverse of the correlation, which
measures the degree of association between two neurons taking into account the
state of all other neurons. Under technical assumptions, the inverse
correlation can be interpreted as a measure of association between directly
connected pairs of neurons and filter out indirectly connected pairs of
neurons. In genetics network, partial correlation have already shown to be
effective way to infer gene regulatory networks
\citep{de2004discovery,Schafer:2005}. This approach is fully described in
\cite{sutera2014simple} and is presented in Algorithm~\ref{alg:inverse-
covariance- pseudocode}. We first describe the pre-processing of the signal,
then explained how to compute the partial  correlation statistics.

Under the simplifying assumption that neurons are on-off units, characterised
by short periods of intense activity, or peaks, and longer periods of
inactivity, the first part of our algorithm consists of cleaning the raw
fluorescence data. More specifically, time-series are processed using standard
signal processing filters in order to : (i) remove noise mainly due to
fluctuations independent of calcium, calcium fluctuations independent of
spiking activity, calcium fluctuations in nearby tissues that have been
mistakenly captured, or simply by the imaging process ; (ii) to account for
fluorescence low decay ; and (iii) to reduce the importance of high global
activity in the network. This data pre-processing is standard  in the field of
signal processing \citep{kaiser1977data, oppenheim1983signals}.
The first step of our pipeline is to smooth the fluorescence signal by applying a low-pass filter for filtering out high frequency noise:
\begin{align}
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\label{eq:weighted-asymetric-median}
\end{align}

@@

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is high global activity in the
network with respect to spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough to be able to detect
interactions, and that it would therefore be preferable to lower their
importance by changing their magnitude appropriately. Additionally, it is well-
known that neurons may also spike because of a high global activity
\citep{stetter2012model}. In such context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such, the signal
output by h is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}, \label{eq:magnify-filter}
\end{align}
whose effect is to magnify the importance of spikes that occur in cases of low
global activity (measured by $\sum_j x^t_j$).

Under the assumption that fluorescence level neuron are jointly Gaussian,
\textit{partial correlation} thus measures conditional dependencies between
variables ; therefore it should naturally only detect direct associations
between neurons and filter out spurious indirect effects as mentioned before.

@@

\begin{algorithm}[ht]
\begin{algorithmic}
 \algsetup{linenodelimiter= }

@@

\end{algorithmic}
\caption{Inverse covariance pseudo-code}
\label{alg:inverse-covariance-pseudocode}
\end{algorithm}
